\plaintitle

### 2016-05-16 paper/paper.tex
Beetje gewerkt aan paper (vooral bubbleview tot hiertoe)
Kan zijn da er wa van bubbleview bij relevante literature mag

\begin{abstract}
    This paper details the conceptualization and implementation of an
    interactive visualization of data about NBA seasons including teams, game
    results and player statistics. The d3.js framework was used to develop this
    visualization.
    %TODO add results
\end{abstract}

@@

@@

\section{Visualization}\label{sec:visualization}
The visualization consists of three parts:
\begin{itemize}
    \item The bubble view: a compact view on the play-offs per season
    \item The statistics (zoom) view: a more detailed view on a statistic of a selected team
    \item The team view: a detailed view on how good or bad a team scores on a specific field position
\end{itemize}

<Explain that this complies with the infovis seeking mantra: Overview, zoom&filter, details on demand
Bubble = overview
statistics= details on demand
Talk about th organization of the 3 views + the choice for a tiled approach
- Each page is a 'tile' that we force with fixed scrolling
- Use of the keyboard is possible (space bar to scroll + arrows, arrows for timeline)
>

All three are discussed in more detail below. The three views are available to
the user on a single webpage. The user start on the bubble view and is able to
navigate to the statistics and team views by scrolling down or clicking on
teams.
%The webpage is built in such a way that it gives a hint to the user at the bottom
%or the top of a page when another view is available. By showing a different 
%background color and an arrow, we invite the visitor to explore other views. The 
%user is also at liberty to scroll through these views. However, a fixed scroll is
%implemeted so users will only end in complete views when moving around. Note that 
%the user also will be guided through all 3 views when performing actions described 
%in the section below.

\subsection{The bubble view}
This view supposes to inform the user of the results of the NBA play-offs per 
season. A circle is used to represent a team, the size its SRS score using a '?'
distribution, a stroke to illustrate it's region or medal won during regional 
competition and curved lines to inform the user which teams played against each 
other. The user has a timeline at its disposal to scroll through time. Note that
the team can be recognized by the logo inside the circle. Our initial desing filled
the circle with the red or blue depending on the region where the teams compete. 
However, the shade of colour was darker if their final ranking was higher in their
region. We've replaced this with the logo's of the teams as we thought this is more
user friendly and the extra information of the ranking doesn't outweight.
<picture of the bubble view compared with double elimination bracket>
A more popular visualization of a play-off is the team double elimination bracket.
Although this is a more common representation, we wanted to create a more compact
version by eliminating the recurring representation of a team. We also considered 
enclosed circles and a sunburst.
<figure of 2 alternatives https://thetuftersblog.wordpress.com/2016/03/07/team-ranking-play-off-infovis/>
Although the enclosed circles is a compact representation as well, we have continued 
with our own design as we believe this view is more original and innovative. We 
didn't encouter a similar view when searching for alternatives to visualize a 
play-off or championship. A new visualization has the advantage that it will 
trigger persons as well.
<Talk about the importance of estethics seen in the lesson>
Our initial bubble view was too compact. It connected teams that played against
each other directly. When searching for alternatives to reflect on our choice we
discovered a similar visualization in a totally different context. 
<https://source.opennews.org/en-US/articles/nyts-512-paths-white-house/>
<comparison of original bubble view with white house paths>
This visualization gives a cleaner view of the competitors for a specific team
compared to our original sketch. Hence we decided to adapt our visualization. We
added an intermediate step between two teams to clearer indicate how teams competed
to become NBA champion.
Next to the static view, the user can interact with the visualization. When he 
hoovers over a circle of a team, only the games played by that team are shown by highliting the 
competitors (their circles) and the curved lines that connect them. This view gets 
fixated when the users clicks on a team.
He/she then automatically gets forwarded to the statistics view where the bubble 
view is still represented in the right upper corner to make sure the user keeps
an overview. The statistics view will be explained in more detail in the next section.
< Hierarchy = sorting = facilitate trends of winners
position have a meaning in the bubble view
when selecting circle => zoom/more detail
Bubble principle of 
- Symmetry (whole = play-off)
- conectedness arc = games played
Pre attentive characteristics
- size of circles
- hue to highligt selections or indicate increase decline arrows next view

We considered a map view, but didn't do it because geographical position has no added value
This visualization helped us discover data quality issues => cut off year 1984
>

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been influenced compared to the year before. The view consists of 3 parts:
\begin{itemize}
    \item Context section: the top part should inform the user of his/her
        context
    \item Arrow section: hhis section illustrates why a statistic changed
    \item Small multiples section: hmall multiples of team statistics and how they evolve over time
\end{itemize}

\subsubsection{Context section}
The context section should inform the user at all times of the current selections 
that have been made. The statistics view will always be based on a specific team
and season that have been selected or that are initated. This should be clear based
on the textual information in the top left. Next to that, in the top right a small 
bubble view is shown. It indicates the team selected and how well it performed in 
the play-offs during that season, as explained in the bubble view section. When changing teams and/or season, this view gets 
updated as well as the textual information.
<Explain that the context supports the short term memory of human beings rfr. Monkey Illusion>

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a chosen statistic for the selected team compared with the season before. This should give a visitor more insight into why a statistic changed over time. The selected team is represented by a circle in the same way as explained in the bubble view. The change for a statistic is represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the influence of players who joined the team (team member current season, not previous season)
    \item Arrow on the right of the circle: indicating how the team inherent changed (team member this season and previous season)
    \item Arrow on the left pointing away from the circle: indicating the influence of players who left the team (team member previous season, not current season)
\end{itemize}
<figure of the statistics view>
The shirts next to the arrows are the shirts of players that had the biggest 
contribution to that arrow. Note that these shirts don't change in size. The box on 
the right enables the user to change between 
statistics. Next to that it informs the user of which statistic has been chosen. When 
the user hits the circle in the middle representing the teams, he is guided 
to the team view. This view will be explained in more detail in the next section.
<filtering: 
- time & team
- based on exploration (small multiples)
- dynamically: changes impact multiple parts>

@@

\subsection{Story telling}
At the beginning of this project we were able to design the 3 separate views 
quite quickly. Our main problem was to integrate these views as a whole. When we created drafts of two views being the bubble view and the statitics view, we suddenly noted patterns. The Golden State Warriors became NBA champion a couple of years after a great uplift in SRS score. This due to a increase inside the team and because of new players joining the team in 2009. In 
the years following, the team kept on attracting talented players and increased the inherent team score (with ups and downs) to become NBA champion in 2015. With the team view, one can even notice that the team gets stronger in center team positions during this period. The visualization however is missing at the moment if this reinforcement of the field positions
is due to the attraction of talented people or by utilizing available talent within the team. Rfr to possible improvements. Being able to visualize this story, enabled us to support visistors in searching their own stories. They are able to explore the data for patters. On top of that, we provide the insight on how the changes were impacted by changes in teamplayers.

>>>>

\begin{abstract}
    In this paper we describe an interactive visualization of the NBA competition over the last 30 years. With three different views, we give the opportunity for the user to explore various kinds of data about NBA seasons including standings, team statistics, transfers and player statistics. The d3.js framework was used to develop this visualization.
    %TODO add results

\end{abstract}

@@

The National Basketball Association (NBA) is the most famous and perhaps the best professional basketball league in the world. Like in every other sports branch, a lot of data and statistics about players, teams, games and seasons are available, mainly in large tables. Despite these tables contains a lot of information, it is very hard if not impossible, to interpret, compare efficiently and to gather insights in it. In our work we try to present this data in a whole different way. Therefore we have focused ourselves on some parts and not the entire available data.

@@

%Related work/References

%http://courses.ischool.berkeley.edu/i247/f05/readings/Baldonado_MultipleViews_AVI00.pdf
%http://lemonchiu.github.io/NBA-Visualization/
%http://peterbeshai.com/buckets/app
%https://source.opennews.org/en-US/articles/nyts-512-paths-white-house/
%http://projects.fivethirtyeight.com/complete-history-of-the-nba/#warriors
%http://www.theguardian.com/news/datablog/interactive/2013/jun/03/premier-league-season-visualised
%http://gizmodo.com/5927503/all-the-major-sport-competitions-since-1903-condensed-in-beautiful-circular-graphics
%http://www.nbaplayoffsbracket.com/2016/index.php
%http://inf.ufrgs.br/~lsguedes/skooth/Publications_&_Works_files/nbavis_final.pdf
%https://en.wikipedia.org/wiki/National_Basketball_Association

\section{Visualization}\label{sec:visualization}
The visualization consists of three parts:
\begin{itemize}
    \item The play-off view: a compact view on the play-offs per season
    \item The statistics (zoom) view: a more detailed view on a statistic of a selected team
    \item The team view: a detailed view on how good or bad a team scores on a specific field position
\end{itemize}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/playoffviewwithcontext}
  \caption{The play-off view without a team selected.}~\label{fig:playoffviewnoteam}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/playoffviewteamselected}
  \caption{The play-off view with a team selected.}~\label{fig:playoffviewteam}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/statisticsview}
  \caption{The statistics view.}~\label{fig:statisticsview}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/teamview}
  \caption{The team view.}~\label{fig:teamview}
\end{figure}

<Explain that this complies with the infovis seeking mantra: Overview, zoom&filter, details on demand
play-off view = overview
statistics view= details on demand
Talk about th organization of the 3 views + the choice for a tiled approach
- Each page is a 'tile' that we force with fixed scrolling
- Use of the keyboard is possible (space bar to scroll + arrows, arrows for timeline)
>

All three are discussed in more detail below. The three views are available to
the user on a single webpage. The user start on the play-off view and is able to
navigate to the statistics and team views by scrolling down or clicking on
teams. A different season can be selected with the left and right arrows or by clicking on the timeline. The timeline is a fixed part of the visualization, hence visible in the three different views.
%The webpage is built in such a way that it gives a hint to the user at the bottom
%or the top of a page when another view is available. By showing a different 
%background color and an arrow, we invite the visitor to explore other views. The 
%user is also at liberty to scroll through these views. However, a fixed scroll is
%implemeted so users will only end in complete views when moving around. Note that 
%the user also will be guided through all 3 views when performing actions described 
%in the section below.

\subsection{The play-off view}
This view informs the user of the roster and end positions in the NBA play-offs for a selected season, in general or for a selected team. The view consists of two parts: 
\begin{itemize}
    \item Context section: the top part informs the user about the selected team and shows a legend of the play-off view.
    \item Bubble section: this section represents the play-offs of the selected season
\end{itemize}
The play-off view is shown in Figure~\ref{fig:playoffviewwithcontext}.

\subsubsection{Context section}
The context section contains a legend that explains the stroke colors used in the bubble section. The stroke is gold, silver or bronze if the team ended respectively first, second or third in the regular competition. Teams that didn't end in a podium place during the regular competition have a red or blue stroke color indicating the team's region, respectively the western or eastern conference. The context section also contains a link to another page (``About'') where the user can get an explanation of the visualization with its different views, can get some background information about the NBA and the terminolgy used, and can check the source of our used data. If the user hovers over or selects a team in the bubble section, the context section shows more information about that team, as shown in Figure~\ref{fig:playoffviewteamselected}. This includes the Simple Rating System score (SRS), the end position in the play-offs and the end position in the regular competition.

\subsubsection{Bubble section}
This section presents the roster and end positions of the play-offs for a selected season. Every team is respresented by a circle with the logo of the team inside it. Our initial desing filled the circle with the logo of the team and colored the circle red or blue depending on the region where the team competed. The shade of colour was darker if their final ranking was higher in their region. According to the Ranking of Perceptual Tasks~\cite{perceptualranking} this is a good property for ordered values like the regular competition ranking. But the combination of the logo's of the teams with the colors made it to heavy and less easy to compare the end ranking of different teams. As an alternative we tried the same view without the logo's, but this is inconvenient because you don't know which team is represented by which circle. Therefore we've chosen to only use the logo's of the teams as we thought this is more user friendly and the extra information of the ranking doesn't outweight. The size of the circle represents the SRS score using an own interpretation of perceptual scaling because the (Flannery) Appearance Compensation is not usable with a domain of both negative and positive values. An exponential function with exponent 1.5 was used. This way the proportion between the area of a bigger and the area of a smaller circle will be greater than the proportion of values the areas represent. The stroke of the circle illustrates the team's region or medal won during the regional competition as explained in the context section. The curved lines between teams represent that those teams have encountered each other in the playoffs. Each row represents an end position in the play-offs. The higher a team is, the further it has come. From the top to the bottom the rows represent the winner of the final (place 1), loser of the final (place 2), losers of the semi-finals (place 3), losers of the quarter-finals (place 4), losers of the 1/8-finales (place 5). The last row represents the teams that did not make it to the play-offs (place 6).  A connection is always made between a team that is in a higher row and a team that is in a lower row. The highest team of the two teams has won the confrontation, and therefor went at least one round further in the play-offs. The two up most teams are the winners of the play-offs in their own conference and played the final. 
A more popular visualization of a play-off is the team double elimination bracket.
Although this is a more common representation , and there are some nice desings like in this NBA 2012 play-offs visualisation~\cite{tournamentladder} where a sunburst visualisation is used, we wanted to create a more compact version by eliminating the recurring representation of a team. We considered enclosed circles, as used in this visualisation of the wimbledon tournament roster~\cite{enclosedcircles}.
Although the enclosed circles is a compact representation as well, we have continued 
with our own design as we believe this view is more original and innovative. Furthermore we are convinced that our view gives a good illustration of the end positions. It's generally assumed that the team that is on top of the ``ladder'' is the winning team. We didn't encouter a similar view when searching for alternatives to visualize a play-off or championship. A new visualization has the advantage that it will trigger users as well.

<Talk about the importance of estethics seen in the lesson>
Our initial play-off view was too heavy. It connected teams that played against
each other directly. When searching for alternatives to reflect on our choice we
discovered a similar visualization in a totally different context. 
<https://source.opennews.org/en-US/articles/nyts-512-paths-white-house/>
<comparison of original play-off view with white house paths>
This visualization gives a cleaner view of the competitors for a specific team
compared to our original sketch. Hence we decided to adapt our visualization. We
added an intermediate step between two teams to clearer indicate how teams competed
to become NBA champion.
Next to the static view, the user can interact with the visualization. When he 
hoovers over a circle of a team, only the games played by that team are shown by highliting the 
competitors (their circles) and the curved lines that connect them. This view gets 
fixated when the users clicks on a team.
He/she then automatically gets forwarded to the statistics view where the play-off 
view is still represented in the right upper corner to make sure the user keeps
an overview. The statistics view will be explained in more detail in the next section.

< Hierarchy = sorting = facilitate trends of winners
position have a meaning in the play-off view
when selecting circle => zoom/more detail
Bubble principle of 
- Symmetry (whole = play-off)
- conectedness arc = games played
Pre attentive characteristics
- size of circles
- hue to highligt selections or indicate increase decline arrows next view

We considered a map view to just show the teams and there regular competition ranking. But this was a view that has no additional value in comparison with the play-off view except for the geographical position, which has no importance for our visualisation. Furthermore we've found a visualisation of the NBA with a map view~\cite{mapviewvisualization} similar to our idea, so we decided to not implement this map view.

This visualization helped us discover data quality issues => cut off year 1984
>

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been influenced in the selected season compared to the season before. The statistics view is showed in Figure~\ref{fig:statisticsview}. The view consists of 3 coordinated parts:
\begin{itemize}
    \item Context section: the top part informs the user of the context of the selected team
    \item Arrow section: this section illustrates why a statistic changed compared to the previous season
    \item Small multiples section: small multiples of team statistics and how they evolve over time
\end{itemize}

\subsubsection{Context section}
The context section informs the user at all times of the current selection of a team  
that have been made in the play-off view. The statistics view will always be based on a specific team and season that have been selected or that are initated. This should be clear based on the textual information in the top left. Next to that, in the top right a small play-off view is shown. It indicates the team selected and how well it performed in  the play-offs during that season, as explained in the play-off view section. When changing teams and/or season, this view gets updated as well as the textual information. <Explain that the context supports the short term memory of human beings rfr. Monkey Illusion>

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a chosen statistic for the selected team compared with the season before. This should give a visitor more insight into why a statistic changed over time. The selected team is represented by a circle in the same way as explained in the play-off view. The change for a statistic is represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the influence of players who joined the team (team member current season, not previous season)
    \item Arrow on the right of the circle: indicating how the team inherent changed (team member this season and previous season)
    \item Arrow on the left pointing away from the circle: indicating the influence of players who left the team (team member previous season, not current season)
\end{itemize}
<figure of the statistics view>
The shirts next to the arrows are the shirts of players that had the biggest 
contribution to that arrow. Note that these shirts don't change in size. The box on 
the right enables the user to change between 
statistics. Next to that it informs the user of which statistic has been chosen. When 
the user hits the circle in the middle representing the teams, he is guided 
to the team view. This view will be explained in more detail in the next section.
<filtering: 
- time & team
- based on exploration (small multiples)
- dynamically: changes impact multiple parts>

@@

\subsection{Story telling}
At the beginning of this project we were able to design the 3 separate views 
quite quickly. Our main problem was to integrate these views as a whole. When we created drafts of two views being the play-off view and the statitics view, we suddenly noted patterns. The Golden State Warriors became NBA champion a couple of years after a great uplift in SRS score. This due to a increase inside the team and because of new players joining the team in 2009. In 
the years following, the team kept on attracting talented players and increased the inherent team score (with ups and downs) to become NBA champion in 2015. With the team view, one can even notice that the team gets stronger in center team positions during this period. The visualization however is missing at the moment if this reinforcement of the field positions
is due to the attraction of talented people or by utilizing available talent within the team. Rfr to possible improvements. Being able to visualize this story, enabled us to support visistors in searching their own stories. They are able to explore the data for patters. On top of that, we provide the insight on how the changes were impacted by changes in teamplayers.
